---
title: "AppliedFinalNeat"
output: pdf_document
date: "2025-03-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Install packages

packages <- c(
  "tidyverse",
  "ggplot2",
  "corrplot",
  "ggcorrplot",
  "caret",
  "data.table",
  "mltools",
  "xgboost",
  "reshape2",
  "viridis",
  "ggfortify",
  "factoextra",
  "cluster",
  "plotly",
  "ggdendro",
  "sportyR",
  "gridExtra"
)

# Only install if not already installed
new_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
if(length(new_packages)) install.packages(new_packages)

# Load from library
library(tidyverse)
library(ggplot2)
library(corrplot)
library(ggcorrplot)
library(caret)
library(data.table)
library(mltools)
library(xgboost)
library(reshape2)
library(viridis)
library(ggfortify)
library(factoextra)
library(cluster)
library(plotly)
library(ggdendro)
library(sportyR)
library(gridExtra)

# Helper Functions


## Barplot
barplot = function(values){
  n = length(values)
  df = data.frame(value = values,  index = 1:n)
  ggplot(df, aes(index, value, fill = value)) + 
    geom_bar(color = "black", stat = "identity")+
    scale_fill_gradient2(low="#619CFF", mid="white", high="#F8766D")+
    theme_bw()
}

## Scatterplot
scatterplot = function(X, cluster = NULL){
  if(is.null(cluster)) cluster = rep(1, nrow(X))
  if(length(unique(cluster)) == 1){
    ggplot()+
      geom_point(X, mapping = aes(hc_x, hc_y)) +
       scale_y_reverse()
  } else {
    ggplot()+
      geom_point(data.frame(X, cluster = as.factor(cluster)), alpha=0.5, 
                 mapping = aes(hc_x, hc_y, color = cluster, shape=type_of_hit))+
       scale_y_reverse() +
  labs(title = "Hit Coordinates of Batted Balls",
       x = "X",
       y = "Y", 
       caption = "Figure _")

  } 
}

## Heatmap
heatmap = function(A){
  n = nrow(A)
  p = ncol(A)  
  df = data.frame(value = c(A),  i = 1:n, j = rep(1:p, rep(n, p)))
  ggplot(df, aes(j, i, fill = value)) + 
    geom_tile()+
    scale_fill_viridis()+
    scale_y_reverse()+
    theme_void()
}
```

```{r}
# Set seed for reproducibility
set.seed(1013)

# Load in data
data <- read_csv("Statcast_2019.csv")
# Remove deprecated and irrelevant columns
data_cleaned <- data %>% select(-contains("deprecated"), -spin_dir, -umpire, 
                                -game_date, -game_type) %>% 
  mutate(team_at_bat = ifelse(inning_topbot == "Top", "away", "home")) %>% 
  drop_na(release_speed)

```


## Q2: Clustering hit ball locations

Suppose we want to cluster the x,y coordinates of where hit balls land (or are caught) in the field. 

```{r}
# Create type_of_hit column
data_cleaned <- data_cleaned %>% 
  mutate(type_of_hit= case_when(
    events == "sac_fly_double_play" | events == "sac_fly" ~ "sac_fly", 
    events == "sac_bunt_double_play" | events == "sac_bunt" ~ "sac_bunt", 
    events == "home_run" ~ "home_run", 
    events == "single" ~ "single",
    events == "double" ~ "double", 
    events == "triple" ~ "triple", 

  )) %>% 
  mutate(location_x = 2.5 * (hc_x - 125.42),
         location_y = 2.5 * (198.27 - hc_y))

# Save smaller, reduced df 
only_hits <- data_cleaned %>% select(hc_x, hc_y, type_of_hit) %>% drop_na()

# Plot the field
ggplot()+
  geom_point(only_hits, alpha=0.1, mapping = aes(hc_x, hc_y, 
                                                 colour = type_of_hit)) + 
    scale_y_reverse() + 
  labs(title = "Hit Coordinates of Batted Balls",
       x = "X",
       y = "Y") +
  guides(colour = guide_legend(title = "Type of Hit", 
                               override.aes = list(alpha = 1))) 

```

We can see that different types of hit balls are concentrated in different parts of the field. For example, sacrificial bunts (`sac_bunt`) are mostly located near the home plate at the bottom vertex of the diamond. 

Can a clustering algorithm recover these patterns? If we simply were to cluster this data, would we be able to neatly classify different types of hit balls? 

Given the noisy, overlapping, non-linear structure of this data, we elected to employ spectral clustering to see if these patterns can be recovered. 

### Spectral Clustering 

```{r}
set.seed(123)
# Original data was too large -- taking random subsample 
only_hits_reduced <- sample_n(only_hits, size=5000)

ggplot()+
  geom_point(only_hits_reduced, alpha=0.2, mapping = aes(hc_x, hc_y, 
                                                         colour = type_of_hit)) + 
  scale_y_reverse() + 
  labs(title = "Hit Coordinates of Batted Balls",
       x = "X",
       y = "Y", 
       caption = "Figure _") +
  guides(colour = guide_legend(title = "Type of Hit", 
                               override.aes = list(alpha = 1))) + theme_bw()
```


```{r}
# Create matrix of location data
loc_data <- only_hits_reduced %>% select(hc_x, hc_y)
loc_matrix <- as.matrix(loc_data)
loc_matrix <- scale(loc_matrix, scale = T, center = T)
```


```{r}
set.seed(123)
# Perform spectral clustering

# KNN function 
knn = function(values, k){
  index = order(values, decreasing = TRUE)[(1:k) + 1]
  values[-index] = NA
  return(values)
}


# Similarity matrix S
S = exp(-1/2 * as.matrix(dist(loc_matrix)))

# Weighted adjacency matrix W, k=20
W = apply(S, 2, function(x) knn(x, k = 5))
W = (W + t(W)) / 2
W[is.na(W)] = 0

# Compute Laplacian L
D = diag(colSums(W))
L = D - W

# Find eigen decomposition of L
n = nrow(L)
ED = eigen(L)
```

```{r}
set.seed(123)
# Try different number of clusters 
plot_list <- list()  # Initialize an empty list to store the plots

# Try different number of clusters 
for (K in 1:6) {
  # Find last K eigen vectors of L 
  U <- ED$vectors[, (n-K+1):n]
  
  # Run k-means on last K eigen vectors of L 
  KM <- kmeans(U, centers = K, iter.max = 100, nstart = 10)
  
  # Generate and store the scatterplot
  plot_list[[K]] <- scatterplot(only_hits_reduced, KM$cluster) +
    ggtitle(paste("K =", K))  # Optional: add title to each plot
}

# Display plots in a grid (e.g., 2 rows, 3 columns)
grid.arrange(grobs = plot_list, nrow = 2, ncol = 3)
```

```{r, fig.width=8, fig.height=6}
# Inspect K=6

set.seed(123)
K = 6
U = ED$vectors[, (n-K+1):n]
# Run k-means on last K eigen vectors of L 
KM = kmeans(U, centers = K, iter.max = 100, nstart = 10)
# Print output 

plt <- ggplot()+
      geom_point(data.frame(only_hits_reduced, cluster = as.factor(KM$cluster)), 
                 alpha=0.5, mapping = aes(hc_x, hc_y, color = cluster, shape=type_of_hit))+
       scale_y_reverse()+ 
  labs(title = "Hit Coordinates of Batted Balls",
       x = "X",
       y = "Y", 
       caption = "Figure _") +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) + theme_bw()

plt

```




## Q3: What are the key factors contributing to player performance in different game situations?


To understand how different game conditions affect a player’s performance, we propose implementing PCA to see if specific features (e.g. the type of game, whether it is at home or away) impact bat outcomes like estimated batting average or foul ball percentage. Again, understanding how different game conditions impact player performance may impact training plans ahead of certain types of games. 


```{r}
# Clean data
data_q3 <- data_cleaned %>% select(-player_name, -batter, -pitcher, -des, -sv_id, 
                                   -game_pk, -contains("fielder"),-game_year, 
                                   -on_1b, -on_2b, -on_3b, -pitcher.1, 
                                   -inning_topbot, -...1)


```

Impute missing data:

- Events: event of resulting plate appearance. If there is no event, there was either a ball, or non-terminal foul or strike (not the third). So, can impute "none" for missing variables here. 
- Hit Location: if no hit location, the ball never made it to the field - also could be a hit or strike. Can impute "none" for these vars too. 
- bb-type: the type of batted ball. again, if the player strikes out, there won't be a type. can input none
- hc_x, hx_y: x,y coordinates of hit ball. Again, if ball isn't hit, we can input 0
- hit_distance, launch_speed, launch_angle: characteristics of hit balls. If not hit, can input none
- estimated_ba, estimated_woba, woba_value, woba_denom, babip_value, iso_value, launch_speed angle: can all replace with None 

```{r}
total_cols = ncol(data_q3)
# Find character columns
character_cols = colnames(data_q3[, sapply(data_q3, class) == 'character'])
# Find numeric columns
numeric_cols = colnames(data_q3[, sapply(data_q3, class) == 'numeric'])

# For numeric columns, impute "None" for missing
data_q3_imputed <- data_q3 %>% 
  mutate_at(character_cols, ~replace_na(., "None")) %>% 
  mutate_at(numeric_cols, ~replace_na(., 0))
```

```{r}
#Reduce dataset variables representing game conditions:
# Eg: variables that are established before the batter hits 
data_q3_reduced <- data_q3_imputed %>% select(pitch_type, 
                                              release_speed, 
                                              release_pos_x, 
                                              release_pos_y,
                                              release_pos_z,
                                              zone, 
                                              stand, 
                                              p_throws, 
                                              balls, 
                                              strikes, 
                                              pfx_x, 
                                              pfx_z, 
                                              plate_x, 
                                              plate_z, 
                                              outs_when_up, 
                                              inning, 
                                              vx0, 
                                              vy0, 
                                              vz0, 
                                              ax, 
                                              ay, 
                                              az, 
                                              sz_top, 
                                              sz_bot, 
                                              effective_speed, 
                                              release_spin_rate, 
                                              release_extension, 
                                              release_pos_y, 
                                              launch_speed_angle, 
                                              at_bat_number, 
                                              pitch_number, 
                                              pitch_name, 
                                              home_score, 
                                              away_score, 
                                              bat_score, 
                                              fld_score, 
                                              if_fielding_alignment, 
                                              of_fielding_alignment,
                                              spin_axis, 
                                              team_at_bat, 
                                              woba_value)

X <- data_q3_reduced %>% select(-woba_value)
y <- data_q3_reduced %>% select(woba_value)

```

```{r}
# Find character columns
character_cols = colnames(X[, sapply(X, class) == 'character'])
numeric_cols = colnames(X[, sapply(X, class) == 'numeric'])

# # Scale numeric cols
X_numeric <- X[numeric_cols]
X_normalized <- scale(X_numeric)
y_normalized <- scale(y)


# One-hot encode categorical cols
X_char <- X[character_cols] 
X_char <- lapply(X_char , factor)
X_char <- as.data.table(X_char)
X_onehot <- one_hot(X_char)

# bind back together
X_full <- cbind(as.data.frame(X_numeric), as.data.frame(X_onehot))
```


```{r}
# Calculate the correlation matrix
cor_matrix <- cor(X_numeric)
# Create a basic correlation heatmap using corrplot
corrplot(cor_matrix, diag = FALSE, method = "color", type="upper", tl.cex=0.5,
        mar=c(1,1,2.5,1))
title(main="Correlation Heatmap of Numeric Variables", sub="Figure _")

```





### Run PCA

```{r}
# Run PCA and find scree plot 

PCA = prcomp(X_full, center = TRUE, scale.=FALSE)
summ = summary(PCA)
data.frame(summ$importance)
ve = summ$importance[2,]
barplot(ve)+
  labs(title="Scree Plot of PCA", caption="Figure _") +
  xlab("Principal Component")+
  ylab("Variance Explained")+
  ylim(0, 1)
```





```{r}
sum(ve[1:2])

# First 2 PCs explain 99.5% of the data
```



### Build model 

```{r}
set.seed(1013)
q3_pcs <- PCA$x[,1:2]
q3_pcs <- as.data.frame(cbind(q3_pcs, y_normalized))


trainIndex <- createDataPartition(q3_pcs$woba_value, p = 0.8, list = FALSE)
train_data <- q3_pcs[trainIndex, ]
test_data <- q3_pcs[-trainIndex, ]

# Convert data to matrix format (XGBoost requires matrices)
train_matrix <- as.matrix(train_data[, 1:2])
test_matrix <- as.matrix(test_data[, 1:2])
train_label <- train_data$woba_value
test_label <- test_data$woba_value

```

```{r}

# Convert data to XGBoost matrix format
dtrain <- xgb.DMatrix(
  data = as.matrix(train_data[, -which(names(train_data) == "woba_value")]), 
                      label = train_data$woba_value)




# Define XGBoost parameters
params <- list(
  objective = "reg:squarederror",  # Regression task
  eta = 0.1,                       # Learning rate
  max_depth = 6,                   # Tree depth
  subsample = 0.5                 # Row sampling

)

# Perform 5-fold cross-validation
cv_results <- xgb.cv(
  params = params,
  data = dtrain,
  nrounds = 100,        # Number of boosting rounds
  nfold = 5,            # 5-fold CV
  metrics = "rmse",     # Evaluate using RMSE
  early_stopping_rounds = 10,  # Stop early if no improvement
  verbose = FALSE
)

# Print best iteration based on CV results
best_nrounds <- cv_results$best_iteration
cat("Best nrounds:", best_nrounds, "\n")

# Train final model with best nrounds
xgb_model <- xgboost(
  params = params,
  data = dtrain,
  nrounds = best_nrounds,  # Use best number of rounds
  verbose = FALSE
)


```
```{r}
test_pred <- predict(xgb_model, test_matrix)

test_data <- tibble(pred = test_pred, actual = test_label)
# Plot predictions vs actual bike rental count
ggplot(test_data, aes(x = pred, y = actual)) + 
  geom_point() + 
  geom_abline()

# Calculate RMSE
test_data %>%
  mutate(residuals = actual - pred) %>%
  summarize(rmse = sqrt(mean(residuals^2)))
```

```{r}
# Find range of the outcomes to see if RMSE is decent 
range(test_label)
```



```{r}
# Get feature importance from XGBoost model
importance_matrix <- xgb.importance(model = xgb_model)
print(importance_matrix)

# Plot feature importance
xgb.plot.importance(importance_matrix)
```

```{r}
# Get PCA loadings
loadings <- PCA$rotation[, 1:2]  # Loadings for first 2 PCs

# Find top contributing original variables for the most important PCs
top_pcs <- importance_matrix$Feature[1:2]  # Top 5 PCs from XGBoost
for (pc in top_pcs) {
  cat("\nTop contributing variables for", pc, ":\n")
  print(sort(abs(loadings[, pc]), decreasing = TRUE)[1:5])  
}
```

```{r}

# Create a list to store the top contributing variables for each PC
top_pcs <- importance_matrix$Feature[1:2]  # Top 2 PCs from XGBoost
bubble_data <- list()

for (pc in top_pcs) {
  top_contrib <- sort(abs(loadings[, pc]), decreasing = TRUE)[1:5]  
  top_contrib_df <- data.frame(
    Variable = names(top_contrib),
    Contribution = top_contrib,
    PC = rep(pc, length(top_contrib)),
    Size = abs(top_contrib) * 10
  )
  bubble_data[[pc]] <- top_contrib_df
}

# Combine all data frames into one
bubble_data_all <- do.call(rbind, bubble_data)
bubble_data_all <- bubble_data_all  %>% 
  mutate(vartext = case_when(
    Variable == "spin_axis" ~ "Pitch Spin Axis", 
    Variable == "ax" ~ "Pitch Acceleration, X",
    Variable == "ay" ~ "Pitch Acceleration, Y",
    Variable == "az" ~ "Pitch Acceleration, Z",
    Variable == "vy0" ~ "Pitch Velocity, Y",
    Variable == "effective_speed" ~ "Pitch Speed from Extension",
    Variable == "release_spin_rate" ~ "Pitch Spin Rate"
    
  ))

# Create a bubble plot
ggplot(bubble_data_all, aes(x = reorder(vartext, Contribution), 
                            y = PC, size = Size, color = PC)) +
  geom_point(alpha = 0.6) +  
  scale_size_continuous(range = c(3, 10)) +  #
  labs(title = "Bubble Plot of Top Contributing Variables for PCs", 
       x = "Variable", y = "Principal Component", size = "Contribution", 
       color = "PC", 
       caption = "Figure _") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```









